---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- # GPHS -->

<!-- badges: start -->
<!-- badges: end -->

## GPHS: Hyperparameter selection in Gaussian Process models

The R package 'GPHS' provides functions for MCMC update of the hyperparameters in covariance function in Gaussian process models. It includes the Elliptical Slice Sampling algorithm as discussed in [Nishihara, Murray and Adams (2014)](https://www.cs.princeton.edu/~rpa/pubs/nishihara2014generalized.pdf) and the Surrogate data model Metropolis-Hastings algorithm as discussed in [Murray et. al (2010)](https://arxiv.org/abs/1001.0175). It also includes a function to provide the user with a comparison of the procedures using graphical tools.

### Installation


You can install the most recent version of 'GPHS' package from [GitHub](https://github.com/niladrik/GPHS) using the following commands:
```{r, eval = FALSE}
# Plain installation
devtools::install_github("niladrik/GPHS") # gphs package
# For installation with vignette
devtools::install_github("niladrik/GPHS", build_vignettes = TRUE)
```

### Remainder of work
In the remainder of the semester, I plan to complete the function for surrogate data model and that for graphical comparison of the algorithms. I will include compatibility checks for each functions and then test for the correctness. Also, I will have to write a vignette and include examples for the functions.
